A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-08-26 masters_thesis/si_paper/reviewer_response/response.tex

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}

\begin{document}

We apologize for the long delay in responding to your feedback.  This was due
to our extensive modifications and improvements to the paper.  We have run over
three processor years worth of additional experiments, greatly expanding the
results section, and we restructured the middle section of the paper, hopefully
improving readability.  Specific comments from the reviewers, with our response
to each comment, are found below.

@@

\section*{Associate Editor Comments}

\subsection*{AE.1}

Reviewer comments: The quality of the article will be improved if you include a
review of existing parallelization methods for PSO, to produce a kind of
taxonomy. Then, also include comparisons with these approaches. You currently
compare only with a naive approach and adaptations of your approach.

Response: We added a Related Work section (now section 3), describing previous
approaches at parallelizing PSO.  We also compared to all applicable methods
that we described.

Prior algorithmic research is divided into two camps.  Either this research
focuses on implementing PSO in specific parallel architectures, or they adapt
the algorithm to be better suited to parallel computation.  Work on adapting
the algorithm to parallel computation (where this paper fits) has focused on
functions with very short evaluation times.  We focus on long-running
functions, such as are found in prior application papers, but requires very
different methods than those for functions with short evaluation times.  Thus
most prior algorithmic advances are not very successful in our domain, as we
show in the paper.  The sections to refer to on this subject are section 3,
section 6.5, and section 7.4.

@@

Response: We restructured most of our results section, and now include
statistical significance in the form of bolded entries in a table.  This is
noted in the paper.

\subsection*{AE.3}

Reviewer comments: It will also be good if you do make use of problems where
parallelization has benefit. For the functions that you use, there is not real
not for parallelization. Linked to this, when will parallelization be
beneficial?

Response: We added experiments on a real-world function that takes longer to
evaluate (fitting a model to a large amount of data), providing more motivation
for the parallelization of PSO (those results are shown in section 7.4).  We
also added some discussion of these issues in the Experimental Setup section.
Section 6.1 discusses when the parallel architecture that we propose is useful
in contrast to other architectures, and section 6.3 discusses our use of
benchmark functions to evaluate our methods.

@@

\subsection*{AE.4}

Reviewer comments: Please re-organize section 3: You start with a very long
theoretical discussion, and only then get to section 3.1.

Response: Good point.  We also realized that we tried to cram too much into a
single section, so we split the two main parts of it into two separate
sections.  There is now a section devoted entirely to the mathematics of the
method (Section 4) and a separate section describing its implementation
(which we moved to the appendix, as we thought it was important, but
distracting from the main discussion).  We also put subsections in the new
Section 4 to guide the reader a little better through the section, and gave an
intuitive description of speculative evaluation before jumping into the math.

@@

\section*{Reviewer 1 Comments}

\subsection*{1.1}

Reviewer comments: I'm a bit skeptical about the actual \emph{practical}
significance of the proposed method, [because it is difficult to know for any
given problem what the right swarm size is]. To be fair, this criticism applies
to most contributions focused on algorithms rather than on problems.

Response: This is a fair criticism, as it is true that experimentation with a
problem is required before one can know how best to solve it.  We added two
paragraphs to the end of section 2 to further make this point, making more
prominent reference to the No Free Lunch theorems.  We also added experiments
on a real-world problem, which is found in section 7.4

@@

\subsection*{1.2}

Reviewer comments: The organization and presentation of the ideas can be
greatly improved. The authors mix results and conclusions with the description
of the approach. This doesn't help the goal of the authors. A reader, cannot
appreciate the significance of some result or conclusion if he/she hasn't
understood what the authors are talking about. I recommend the authors to
restructure the whole paper but in particular Sections 3 and 4.

Response: We substantially reorganized Section 3, as noted above under AE.4,
and we added more subsections throughout the paper, helping to guide the reader
through the ideas.  We removed most of the forward pointers in section 4,
focusing simply on describing the methods and leaving the comparison to the
results section, as you suggested.  We believe that the paper is now much more
clear.

@@

\subsection*{1.3}

Reviewer comments: The notation used throughout the paper is confusing. In the
PSO literature, it is common to use x as a particles current position and some
other letters to denote a particle's personal best position and a particle's
neighborhood's best position. I know that the authors are using superscripts to
distinguish between these concepts but they don't really help.

Response: We changed the x to a b when referring to best positions, while still
using the N and P superscripts to differentiate between which best is being
referred to.  We hope the change in notation is helpful.

@@

\subsection*{1.4}

Reviewer comments: Using ``rounds of evaluations'' is confusing, and results
should be reported in terms of wall clock time.

Response: We realize that ``rounds of evaluations'' is not a commonly used
term, so we changed ``rounds of evaluations'' to ``time steps'' in hopes of
better conveying our meaning.  We also added a section in the experimental
setup that discusses the issue of wall clock time (section 6.2), saying clearly
that time steps (or rounds of evaluations, as we used to call them), are
directly proportional to wall clock time.

@@

\subsection*{1.5}

Reviewer comments: Since the authors are using Herrera et al.'s proposed
benchmarks and since they are using a large computing cluster, I don't see why
they can't use the same experimental protocol proposed by Herrera et al., which
is focused on large scale optimization. That protocol would better justify a
parallel algorithm because a 1000-dimensional problem is intuitively more
difficult than the 20-dimensional problems the authors are currently using. As
a side product, the authors can actually compare the results of the pruned
versions (which in my eyes, are new PSO variants) with those of other
state-of-the-art algorithms.

Response: We ran quite a few more experiments to bring our results closer to
the experimental protocol of Herrera et al.  Their protocol wasn't entirely
applicable, as they suggested running a specific number of function evaluations
instead of iterations (which, as we state in section 6.2, is the proper
comparison for parallel algorithms on long running functions).  But we added
experiments for 50 dimensional functions and 500 dimensional functions.

We also see your point in saying that our relaxations to speculative evaluation
are new PSO variants, and we fixed our text in places to better address that
point.  In particular, we changed the paragraph in the conclusion that deals
with this issue (``What we have presented\ldots''), clearly stating that our
methods do in fact change the behavior of PSO, but that they do so in a way
that is applicable to almost all PSO variants.  Thus a fair comparison is to
pick a particular variant and compare parallelization techniques with it.  It
made sense to pick the standard PSO for this comparison.  To further show this
point, we also ran a simple comparison of our parallelization techniques using
a different PSO variant, Bare Bones PSO.  Those results are in section 7.5.

@@

\subsection*{1.6}

Reviewer comments: In the abstract, it is not clear the meaning of
``speculative''. A reader without knowledge about parallel computing may not
understand what the authors mean.

Response: We had a paragraph explaining this in the introduction, though the
analogy to processors lacked an explicit connection to the word
``speculative.''  We added text at the end of that paragraph to address this
issue.

@@

Response: We added the citation as suggested.

\subsection*{1.8}

Reviewer comments: What serves as the motivation for the authors's work is
insufficiently discussed.  The authors say that adding more particles leads to
``diminishing returns''. This may be true when dealing with constant size
populations.  However, a recent trend is to vary the population size over time.
In fact, it has been shown that adding particles \emph{over time} may actually
solve the problem the authors refer to. Some references are ``Incremental
Social Learning in Particle Swarms'' by Montes de Oca and others published in
IEEE SMCB and ``Efficient Population Utilization Strategy for Particle Swarm
Optimizer'' by Hsieh and others in the same journal.

Response: We added a paragraph in our new related work section discussing these
methods (the last paragraph of the section).  We also added a paragraph to
future work discussing how these methods are particularly amenable to
speculative evaluation when parallelized (the second paragraph of the section).

There is also now a graph in section 7.1.4 that clearly shows what we mean.  It
is Figure 10, showing what happens when processors are added with standard
parallelizations versus with our methods.

@@

\subsection*{1.9}

Reviewer comments: In the first paragraph of Section 3, the authors say that
adding particles may not help past a certain swarm size. This is true if the
computation time allocated to the optimization process remains fixed. However,
in some problems, it is the solution quality what matters. In that case, adding
particles \emph{and} computational budget may indeed be the best thing to do.

Response: This is very much related to the previous item, and so the response
to that item is relevant here.  We also added some text to the introduction
(paragraph beginning with ``In this paper we consider\ldots'') to better
explain our intent; that is, given a fixed computational budget, how can
processors be used most efficiently?  The text we added in this and the
previous item addresses this question.

@@

Response: We fixed section 3 by splitting it into two sections and reorganizing
some parts, making the examples that were there more clear.  Hopefully the
section is now satisfactory.

\subsection*{1.12}

Reviewer comments: Isn't pruning against the idea of better using the available
computing resources? If one reduces the number of evaluated speculative
children, then fewer CPUs are going to be used.

Response: Pruning allows for the use of a larger swarm with the same number of
processors.  We realized that this was never explicitly stated in the text, so
we added a few sentences to the description of pruning (now section 5.2) to
make it clear.  The text added to the introduction explaining our purpose of
fully utilizing resources also addresses this issue, though somewhat
indirectly.

@@

Response: We were referring to the use of ``stagnation'' in analyses of the
sampling distribution of PSO (such as by Poli).  Our results show that the
``lack of improvement'' to which you refer does not in fact happen very often.
It is just that the scale of the improvements becomes increasingly smaller,
sometimes beyond machine precision.  We appreciate your distinction, however.
Such a definition of stagnation must then incorporate a minimum velocity, which
is another function-dependent parameter to specify.  We decided that the
paragraph dealing with this issue was at best tangential to the focus of our
paper and that it unnecessarily lengthened an already lengthy discussion.  We
removed the paragraph from our text.

\subsection*{1.14}

Reviewer comments: All figures are not clear. The authors should find line
styles that are more distinguishable.

Response: We had Gnuplot render the figures in gray mode, using line styles
that are easier to distinguish in black and white.  Hopefully that solves the
problem.  That does lose color in the figures, though.  If color is desirable,
we can find line styles and colors that are both acceptable.  We also put most
of our results into tables, so the reader does not have to rely as much on
interpreting figures.

@@

Response: You seem to have been conflating two separate issues, and we changed
our text to help future readers not run into the same problem.  When we said
that our method ``can be easily extended to arbitrary topologies,'' we were
referring specifically to the math; we only showed one case, to simplify our
notation.  In fact our implementation handles arbitrary topologies.  One thing
that we realized, however, was that with pruning one can use arbitrarily dense
topologies without increasing the number of speculative evaluations at all.  We
thus ran more experiments to test the use of these topologies with pruned
speculative methods, and we now report those results in the paper.

\subsection*{2.3}

Reviewer comments: My personal experience with the PSO versions used in the
paper suggests that, in large dimensions, it is preferrable to use larger
swarms for less iterations rather than small swarms for more iterations. This
is in contrast to the argued behavior of PSO for the Sphere function (Section
1).  Did you observe the same behavior in other functions?  Could you provide
some evidence on this because, due to the swarm's diversity loss, it seems
somehow counterintuitive.

Response: The citation given in the paper (McNabb et al. CEC 09) gives an
interesting discussion on these points.  The functions explored there are
limited, but the paper shows that both Sphere and Griewank exhibit this
behavior, while Rastrigin does not (the relevant figures in that paper are 2, 4
and 6).  As those results are previously published and cited in our paper, we
felt it unnecessary to reduplicate the figures, though we make mention of the
conclusions.  However, our own figures also answer your question.  The fact
that our methods outperform the standard parallelization is in itself evidence
for this claim.  Recall that when our figures say SEPSO, that is equivalent to
running PSO with a swarm size one eighth the size, doing twice as many
iterations.  The direct answer to your question is in what is now Figure 10
(which we added to the paper).  We show that adding particles beyond 240 for a
20 dimensional Griewank function provides no appreciable benefit, while using
speculative methods gives dramatic improvements.

### 2011-08-26 masters_thesis/si_paper/paper.tex
Final improvements from reviewer feedback; revision submitted to SI

However, given this new parallelization model we can relax the requirement of
exactly reproducing PSO in an attempt to produce better results.  We present
several such relaxations, including keeping the best speculative position
evaluated instead of the one corresponding to the standard behavior of PSO, and
speculating several iterations ahead instead of just one.  We show that these
methods dramatically improve the performance of parallel PSO in many cases.

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\citep{parsopoulos-2004-parallel-vector-evaluated-pso} and asynchronous
algorithms \citep{mostaghim-2006-multi-objective-pso-on-grids}.  Parallelizing
the evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

@@

Asynchronous parallel
PSO~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso} is a modification to the standard algorithm
which removes the synchronization point at the end of each iteration.  Instead,
particles iterate independently and communicate asynchronously.  In a typical
master-slave implementation of asynchronous parallel PSO, the master updates
each particle's personal best, neighborhood best, velocity, and position
immediately after receiving the function value from the slave processor.  Since
this update occurs while other particles are still being evaluated, it may use
information from the previous iteration for some
neighbors.\footnote{Asynchronous parallel PSO has been compared to the
``asynchronous updates'' variant of serial
PSO~\citep{koh-2006-parallel-asynchronous-pso}.  However, serial PSO with
asynchronous updates differs from standard PSO in that particles use newer
information, but asynchronous parallel PSO differs from standard PSO in that
particles use older information.} In a partially asynchronous implementation,
particles might wait for some but not all neighbors to complete before
proceeding~%
\citep{scriven-2008-asynchronous-pso-in-unreliable-distributed-environments}.
In some master-slave implementations, particles never get more than one
iteration ahead of
others~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.  However, in a fully distributed
implementation, particles might never wait for information, and one particle
could complete many more iterations than another
particle~\citep{scriven-2008-distributed-pso-using-peer-to-peer}.  The main
effect of asynchronous evaluation is that processors spend less time
idle---this trait is particularly valuable when processors are heterogeneous or
function evaluation times are
varied~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.  Asynchronous parallel PSO behaves
differently than the standard algorithm and may even produce different results
between runs.  Most reports conclude that asynchronous communication produces
similar numerical results to the standard algorithm, but the question has not
yet been thoroughly
addressed~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.

@@

\subsection{Branch Statistics}
\label{sec:statistics}

@@

\subsection{Recovering from Pruning Too Much}
\label{sec:wrong}

@@

Another possibility is to pick the best child, as described in \sec{pickbest}.
This ignores the fact that the branch might have been wrong; it does not
matter, because we simply accept the child that had the best value.  In most of
the experiments that we ran, it turned out that picking the best child
performed better than keeping the particle back an iteration.

@@

We ran experiments to compare our speculative parallelization of PSO to the
standard parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation.  The exact evaluation time at
which this architecture becomes reasonable depends on the amount of
communication overhead in the parallel implementation and the number of
particles in the swarm.  For our implementation we found that time to be around
100 milliseconds for swarms of 50 particles.  When the swarm size increases,
the minimum evaluation time at which this architecture should be used
decreases.

@@

To provide additional evidence that performance on benchmark functions roughly
corresponds to performance on real-world problems, we also tested our
parallelization methods on a typical research problem, that of fitting a model
to a large quantity of data.  We generated 10,000 data points from a radial
basis function network with 10 bases and some added Gaussian noise.  We then
fit a radial basis function network to the data using PSO.  This amounted to a
30 dimensional function to optimize, with a function evaluation time on the
order of two seconds.  We refer to this function as ``the model fitting
problem.''

@@

The parallelization techniques we compare are the standard parallelization
(here labeled Standard simply for ease of reference), our original speculative
approach (recall that we refer to this as SEPSO), and the four relaxations of
SEPSO discussed in \sec{relax}.  In presenting our results, we call the
approach developed in \sec{pickbest} Pick Best.  The methods described in
\sec{pruning} through \sec{wrong} are called Pick Best Pruned and Social
Promotion Pruned, and the method in \sec{manyiters} we call Many Iterations.

@@

@@

It is generally recommended to use the Ring topology when optimizing the
Griewank function, as Complete is prone to premature convergence on a local
optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  The global optimum
of Griewank has a value of 0.  When most trials reach the global optimum but a
few get stuck, the resultant ``average value'' graph has a flat line that is
misleading.  Thus instead we present plots showing the percent of trials that
have passed some threshold at each time step, as is common practice with these
functions~\citep{mendes-2004-population-topologies-in-pso}.  The threshold we
chose for this case was $10^{-6}$, as that value is below any local optima and
the swarm always successfully reaches 0 once is passes that point.

@@

At 500 dimensions the performance of constricted PSO on benchmark functions
becomes rather dismal.  In almost all cases, Standard fails to make any
significant progress.  Our search for an explanation seems to say that the
space is too large for particles to converge to the same point, so all updates
to the best position found come from a single particle wandering in the space.
Every time that single particle finds a new best position, its velocity
contracts, so the particle is unable to make significant progress on its own.
In constricted PSO it is necessary to have a collection of particles exploring
a promising location to keep the particles' velocities from contracting
prematurely, and in 500 dimensions the space is too large to get the collection
of particles to the same location.

@@

One could argue that a fairer comparison would have modified the PSO motions
equations so that individual particles did not contract their velocity so
quickly when finding good locations.  However, we used the same motion
equations for both Standard and Many Iterations; changing the motion equations
to help Standard would also help Many Iterations, and we would expect to see
results that are at least as compelling as those from the 50 dimensional
benchmark functions.

@@

In order to demonstrate that our methods are applicable across PSO variants, we
ran experiments with Bare Bones PSO.  For this simple experiment we use the
Sphere function with 20 dimensions, taking the best Standard algorithm and the
best speculative algorithm on that function, which in this case is Standard
Complete and Pick Best Pruned Complete, as shown in \sec{sphere20dims}.
\fig{bare-bones} shows the results of running each of those algorithms with the
Bare Bones PSO equations.  We see a decrease in time to completion between
Standard and speculative parallelizations of 40\%.  While this is not quite as
large a speed up as the 60\% seen with constricted PSO, it is still a
significant decrease in time.  The key point of this experiment is that our
methods still show improvements over Standard parallelizations in PSO variants
other than the standard constricted PSO.  Our intent is not to give an
exhaustive listing of PSO variants and compare parallelization strategies with
each one; instead we offer evidence supporting our claim that speculative
evaluation should see similar increases in performance across most or all PSO
variants.

@@

We have detailed how to perform speculative evaluation in PSO in several
different parallel architectures.  Using this methodology, the behavior of the
original PSO algorithm can either be exactly reproduced, two iterations at a
time, or the behavior can be modified in order to improve performance.  While
exactly reproducing PSO sometimes uses too many extra processors to be useful,
when we allow ourselves some freedom with the algorithm we see great
improvements over previous methods.  We have shown results that conclusively
demonstrate the superiority of our techniques for several functions over the
standard practice of adding particles to the swarm when extra processors are
available.

What we have presented is not a new variant of PSO.  We presented a new
parallelization technique, so we compared parallelization strategies on the
same algorithm, the original PSO.  Our most promising techniques do change the
behavior of the PSO algorithm slightly, as do some other previously proposed
parallelization techniques, such as asynchronous PSO.  However, our methods are
applicable to almost all PSO variants, and so a comparison using the same
variant for each of the parallelization techniques tested is justified.

We have given five different possible approaches to speculative evaluation,
each of which has different properties.  These approaches perform differently
on different functions and at different swarm sizes, as would be expected by
the No Free Lunch Theorem for
Optimization~\citep{wolpert-1997-nfl-for-optimization}.  We have given a brief
evaluation of the premature convergence properties of these methods on
deceptive functions when a smaller number of processors is available.  We have
also shown evidence that when many processors are available by far the best
thing to do in most cases is to speculate as far ahead as the extra processors
allow.

Though our methods show great improvements on some functions, they do not work
for all functions.  As is commonly known, in PSO there is a trade-off between
exploration and exploitation.  Some functions need only minimal exploration,
and some never seem to have enough.  Increasing the swarm size is a natural way
to increase exploration in a parallel environment.  However, once ``enough''
exploration has been reached for any particular function, adding additional
particles adds only incremental benefits.  At this point, a better use of the
additional processors, as we have shown, is to perform some amount of
speculative evaluation.

@@

\section*{Appendix A---Implementing Speculative Evaluation}
\label{sec:implementation}

@@

\subsection*{Terminology}

@@

\subsection*{Centralized Algorithms}

@@

\subsection*{Distributed Algorithms}

@@

\subsection*{Dynamic Topologies}

@@

\section*{Appendix B---Alternate form of message passing}

>>>>

However, given this new parallelization model we can relax the requirement of
exactly reproducing PSO in an attempt to produce better results.  We present
several such relaxations, including keeping the best speculative position
evaluated instead of the one corresponding to the standard behavior of PSO, and
speculating several iterations ahead instead of just one.  We show that these
methods dramatically improve the performance of parallel PSO in many cases,
giving speed ups of up to six compared to previous parallelization techniques.

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\citep{parsopoulos-2004-parallel-vector-evaluated-pso} and asynchronous
algorithms \citep{mostaghim-2006-multi-objective-pso-on-grids}.  Parallelizing
the evaluation of the objective function can also be done in some cases using
standard techniques \citep{grama-2003-intro-to-parallel-computing}, though that
is not an adaption of the PSO algorithm itself and thus is not the focus of
this paper.

@@

Asynchronous parallel particle swarm
optimization~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso} is a modification to the standard algorithm
which removes the synchronization point at the end of each iteration.  Instead,
particles iterate independently and communicate asynchronously.  In a typical
master-slave implementation of asynchronous parallel PSO, the master updates
each particle's personal best, neighborhood best, velocity, and position
immediately after receiving the function value from the slave processor.  Since
this update occurs while other particles are still being evaluated, it may use
information from the previous iteration for some
neighbors.\footnote{Asynchronous parallel PSO has been compared to the
``asynchronous updates'' variant of serial
PSO~\citep{koh-2006-parallel-asynchronous-pso}.  However, serial PSO with
asynchronous updates differs from standard PSO in that particles use newer
information, but asynchronous parallel PSO differs from standard PSO in that
particles use older information.} In a partially asynchronous implementation,
particles might wait for some but not all neighbors to complete before
proceeding~%
\citep{scriven-2008-asynchronous-pso-in-unreliable-distributed-environments}.
In some master-slave implementations, particles never get more than one
iteration ahead of
others~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.  However, in a fully distributed
implementation, particles might never wait for information, and one particle
could complete many more iterations than another
particle~\citep{scriven-2008-distributed-pso-using-peer-to-peer}.  The main
effect of asynchronous evaluation is that processors spend less time
idle---this trait is particularly valuable when processors are heterogeneous or
function evaluation times are
varied~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.  Asynchronous parallel PSO behaves
differently than the standard algorithm and may even produce different results
between runs.  Most reports conclude that asynchronous communication produces
similar numerical results to the standard algorithm, but the question has not
yet been thoroughly
addressed~\citep{venter-2005-parallel-pso-asynchronous-evaluations,
koh-2006-parallel-asynchronous-pso}.

@@

\subsubsection{Branch Statistics}
\label{sec:statistics}

@@

\subsubsection{Recovering from Pruning Too Much}
\label{sec:wrong}

@@

Another possibility is to pick the best child, as described in \sec{pickbest}.
This ignores the fact that the branch might have been wrong; it does not
matter, because we simply accept the child that had the best value.

@@

\subsection{Parallel Architecture}

At each iteration of the algorithms, we use one processor to perform one
function evaluation.  The exact evaluation time at which this architecture
becomes reasonable depends on the amount of communication overhead in the
parallel implementation and the number of particles in the swarm.  For our
implementation we found that time to be around 100 milliseconds for swarms of
50 particles.  When the swarm size increases, the minimum evaluation time at
which this architecture should be used decreases.

A general rule of thumb for when this architecture is useful is when the
combined function evaluation time is greater than the communication overhead
for a single iteration; i.e., $f*p > o+f$, where $f$ is the function evaluation
time, $p$ is the desired number of particles (in our case $p$ is equal to the
number of available processors), and $o$ is the amount of overhead per
iteration.  The amount of overhead per iteration will vary based on the
communication topology and the number of particles, so this is only an
approximation, but it can provide rough estimates of function evaluation times
at which this parallel architecture should be used.

@@

To provide additional evidence that performance on benchmark functions
corresponds to performance on real-world problems, we also tested our
parallelization methods on a typical research problem, that of fitting a model
to a large quantity of data.  We generated 10,000 data points from a radial
basis function network with 10 bases and some added Gaussian noise.  We then
fit a radial basis function network to the data using PSO.  This amounted to a
30 dimensional function to optimize, with a function evaluation time on the
order of two seconds.  We refer to this function as ``the model fitting
problem.''

@@

The parallelization techniques we compare are the standard parallelization
(here labeled Standard simply for ease of reference), our original speculative
approach (recall that we refer to this as SEPSO), and the four relaxations of
SEPSO discussed in \sec{relax}.  In presenting our results, we call the
approach developed in \sec{pickbest} Pick Best.  The methods described in
\sec{pruning} are called Pick Best Pruned and Social Promotion Pruned, and the
method in \sec{manyiters} we call Many Iterations.

@@

Finally, to show that our methods are orthogonal to traditional PSO variants,
we show results of parallelizing a variation of the PSO motion equations, Bare
Bones PSO~\citep{kennedy-2003-bare-bones-pso}, with both standard
parallelizations and our speculative approaches.  Our intent is not to
exhaustively compare parallelization strategies on all possible PSO variants,
but to show some evidence that speculative approaches are generally applicable
and successful across PSO variants.  These results are shown in
\sec{bare-results}.

@@

It is recommended in the literature to use the Ring topology when optimizing
the Griewank function, as Complete is prone to premature convergence on a local
optimum~\citep{bratton-2007-defining-a-standard-for-pso}.  The global optimum
of Griewank has a value of 0.  When most trials reach the global optimum but a
few get stuck, the resultant ``average value'' graph has a flat line that is
misleading.  Thus instead we present plots showing the percent of trials that
have passed some threshold at each time step, as is common practice with these
functions~\citep{mendes-2004-population-topologies-in-pso}.  The threshold we
chose for this case was $10^{-6}$, as that value is below any local optima and
the swarm always successfully reaches 0 once is passes that point.

@@

At 500 dimensions the performance of constricted PSO on benchmark functions
becomes rather dismal.  In almost all cases, Standard fails to make any
significant progress.  Our search for an explanation seems to say that the
space is too large for particles to converge to the same point (due to
overshooting in some dimension), so all updates to the best position found come
from a single particle wandering in the space.  Every time that single particle
finds a new best position, its velocity contracts, so the particle is unable to
make significant progress on its own.  In constricted PSO it is necessary to
have a collection of particles exploring a promising location to keep the
particles' velocities from contracting prematurely, and in 500 dimensions the
space is too large to get the collection of particles to the same location.

@@

@@

In order to demonstrate that our methods are applicable across PSO variants, we
ran experiments with Bare Bones PSO.  For this simple experiment we use the
Sphere function with 20 dimensions, taking the best Standard algorithm and the
best speculative algorithm on that function, which in this case is Standard
Complete and Pick Best Pruned Complete, as shown in \sec{sphere20dims}.
\fig{bare-bones} shows the results of running each of those algorithms with the
Bare Bones PSO equations.  We see a speed up of 1.6 when using speculative
parallelizations instead of Standard techniques.  While this is not quite as
large a speed up as the 2.5 seen with constricted PSO, it is still a
significant decrease in time.  The key point of this experiment is that our
methods still show improvements over Standard parallelizations in PSO variants
other than the standard constricted PSO.

@@

We have detailed how to perform speculative evaluation in PSO in several
different parallel architectures.  Using this methodology, the behavior of the
original PSO algorithm can either be exactly reproduced, two iterations at a
time, or the behavior can be modified in order to improve performance.  While
exactly reproducing PSO sometimes uses too many extra processors to be useful,
when we allow ourselves some freedom with the algorithm we see great
improvements over previous methods.  We have shown results that conclusively
demonstrate the superiority of our techniques for several functions over the
standard practice of adding particles to the swarm when extra processors are
available, giving speed ups of up to six compared to previous parallelization
techniques.

What we have presented is not a new variant of PSO.  We presented a new
parallelization technique, so we compared parallelization strategies on the
same algorithm, the original PSO.  Our most promising techniques do change the
behavior of the PSO algorithm slightly, as do some other previously proposed
parallelization techniques, such as asynchronous PSO.  However, our methods are
applicable to almost all PSO variants, and so a comparison using the same
variant for each of the parallelization techniques tested is justified.  We
presented results on a separate PSO variant that give evidence for this claim.

We have given five different possible approaches to speculative evaluation,
each of which has different properties.  These approaches perform differently
on different functions and at different swarm sizes, as would be expected by
the No Free Lunch Theorem for
Optimization~\citep{wolpert-1997-nfl-for-optimization}.  We have given a brief
evaluation of the premature convergence properties of these methods on
deceptive functions when a smaller number of processors is available.  We have
also shown evidence that when many processors are available by far the best
thing to do in most cases is to speculate as far ahead as the extra processors
allow.  This point is shown in a single graph in \fig{griewank-all}.  In that
case, the extra processors in standard parallelizations provided \emph{no
appreciable benefit.}  In contrast, speculative methods allow additional
processors to provide arbitrarily high speed ups, as the more processors are
available the farther ahead the algorithm can speculate.

Though our methods show great improvements on some functions, they do not work
for all functions at the swarm sizes we were able to experiment with.  As is
commonly known, in PSO there is a trade-off between exploration and
exploitation.  Some functions need only minimal exploration, and some never
seem to have enough.  Increasing the swarm size is a natural way to increase
exploration in a parallel environment.  However, once ``enough'' exploration
has been reached for any particular function, adding additional particles adds
only incremental benefits.  At this point, a better use of the additional
processors, as we have shown, is to perform some amount of speculative
evaluation.

@@

\appendix

\section{Implementing Speculative Evaluation}
\label{sec:implementation}

@@

\subsection{Terminology}

@@

\subsection{Centralized Algorithms}

@@

\subsection{Distributed Algorithms}

@@

\subsection{Dynamic Topologies}

@@

\section{Alternate form of message passing}
